---
title: "model"
author: "Karin Näsvall"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model recombination rate

Issue	    Tool
**Collinearity**	Correlation matrix + VIF
**Non-normality of predictors**	Spearman correlation, rank-transform, or GAM
**Spatial autocorrelation**	Random effects for genome/chr, or GAM with s(window_mid)
**Non-linear effects**	mgcv::gam() with smooth terms
For highly non-linear relationships (likely for recombination), a GAM (mgcv) is often more appropriate than linear models — start with the correlation/VIF step regardless, then decide between lmer and gam.


1. Pivot wide and explore distributions

```{r data}
library(broom.mixed)
library(dplyr)
library(GGally)
library(ggpubr)
library(gratia)

 # if this complains about dplyr version restart and load dplyr before tidyverse
library(tidyverse)
library(patchwork)

pop_gen_feat <- c("dxy", "Fst", "pi_Mpol", "pi_Mlys", "sites")
repeat_feat <- c("Total_repeats","LTR", "LINE", "DNA", "RC", "Simple_repeat","SINE", "Overlapping", "Low_complexity", "Unknown")
other_features <- c("GC","Genes","eig_scores", "E2_corr", "Recombination_rate")


combined_df_brp <- read_csv("tables/combined_df_incl_brp_rel_pos2026-06-19.csv") %>% filter(!feature %in% pop_gen_feat)

combined_df_brp$break_status <- combined_df_brp$type
combined_df_brp <- combined_df_brp %>% filter(!type %in% c("Complex")) %>% select(-type)


wide_df <- combined_df_brp %>%
  pivot_wider(names_from = feature, values_from = value) 

# Classify each chromosome as colinear (all windows colinear) or rearranged (any non-colinear window)
chr_class <- wide_df %>%
  group_by(genome, chr) %>%
  summarise(chr_category = if (all(event_timing %in% c("colinear", "end", "shared_event_n4", "anc_proxy"), na.rm = TRUE)) "colinear" else "rearranged",
            .groups = "drop")

View(chr_class)

wide_df <- wide_df %>%
  left_join(chr_class, by = c("genome", "chr"))

# remove sex chromosomes from the model
wide_df <- wide_df %>% filter(!chr %in% c("Z", "W", "Z1", "Z2", "W1")) 

wide_df <- wide_df[!(wide_df$genome == "ilMecLysi212" & wide_df$chr == "10"), ]
combined_df_brp <- left_join(combined_df_brp, chr_class)
wide_df
```


## 2. Check distributions, pairwise correlations, and collinearity first

This shows which variables to consider dropping or combining.

```{r}
library(corrplot)

feat_cols_all <- setdiff(names(wide_df), c("chr", "start", "end", "genome","window_mid", "break_status","event_timing", "chr_category", "i.start", "i.end"))

cor_mat <- cor(wide_df[, feat_cols_all], use = "pairwise.complete.obs", method = "spearman")

# Blue–white–red (classic diverging)
col = colorRampPalette(c("royalblue4", "white", "firebrick"))(200)

# clustered correlation matrix with coefficients
pdf("plots/correlation_matrix_rr.pdf", width = 16, height = 16)
corrplot(cor_mat, method = "color", order = "hclust", tl.cex = 0.7,
         addCoef.col = "black", number.cex = 1, 
         col = colorRampPalette(c("#0072B2",  "white", "#D55E00"))(200))
dev.off()

# Pairwise scatterplots with Spearman correlations and distributions on the diagonal
ggpairs(wide_df[, feat_cols_all],
        lower = list(continuous = wrap("points", alpha = 0.1, size = 0.3)),
        diag  = list(continuous = wrap("densityDiag")),
        upper = list(continuous = wrap("cor", method = "spearman")))

ggsave("plots/pairwise_plots_rr.pdf", width = 30, height = 24)



feat_cols <- feat_cols_all[!feat_cols_all %in% c("Total_repeats",  "dist_telomere", "rel_position", "Recombination_rate", "E3", "closest_break_type", "chr_category.x", "chr_category.y" )] # LINEs are driving Total repeats

```

Drop "Total repeat" which is strongly driven by LINEs, distance to telomere which both are almost the same as rel position. 
I also drop recombination rate from the predictors, but keep it as the response variable.


## 3. Remove near-redundant variables (VIF)

Variance Inflation Factor (VIF) measures how much the variance of a regression coefficient is inflated due to collinearity with other predictors.


```{r}

library(car)

# Fit a provisional linear model with all predictors
f <- as.formula(paste("Recombination_rate ~", paste("break_status", paste(feat_cols, collapse = " + "), sep = " + ")))
vif(lm(f, data = wide_df, na.action = na.exclude))

sink("tables/model_VIF.txt")
print(vif(lm(f, data = wide_df, na.action = na.exclude)))
sink()


```


Interpretation:

VIF	Meaning
1	No collinearity
1–5	Moderate, usually acceptable
5–10	High, worth investigating
>10	Severe — predictor is nearly redundant with others

Drop variables with VIF > 10, keeping the one with the clearest biological interpretation.



## 4. Account for spatial autocorrelation

Use a mixed model with genome as a random effect and, if needed, a spatial correlation structure:

```{r}
library(lme4)

m0 <- lmer(as.formula(paste("Recombination_rate ~", paste("break_status", paste(feat_cols, collapse = " + "), sep = " + "), "+ (1 | genome)")),
           data = wide_df, REML = FALSE)


summary(m0)

sink("tables/model_summaries.txt")
cat("\n=== GLM ===\n")
print(summary(m0))
sink()

```

## 5. Standardise predictors for comparability 

```{r}
        
wide_df <- wide_df %>%
  mutate(across(all_of(feat_cols), ~ scale(.)[,1], .names = "{.col}_z"))

feat_cols_sc <- paste0(feat_cols, "_z")

m0_sc <- lmer(as.formula(paste("Recombination_rate ~", paste("break_status", paste(feat_cols_sc, collapse = " + "), sep = " + "), 
              "+ (1 | genome)")),
           data = wide_df, REML = FALSE)

summary(m0_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=m0_sc =\n")
print(summary(m0_sc))
sink()


```

```{r eval_model}


tidy(m0_sc, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Standardised effect on recombination rate", y = NULL) +
  theme_classic()

ggsave(paste("plots/model_effects_rr_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 6)


```

## 6. Model diagnostics and refinement

model diagnostics: check residuals for normality and homoscedasticity, 
and consider alternative models (e.g., GAM) if assumptions are violated.

```{r diagnostics}

pdf(paste("plots/model_diagnostics_rr_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
par(mfrow = c(1, 1))

residuals <- resid(m0_sc)
hist(residuals, main = "Residuals Histogram", xlab = "Residuals", breaks = 30)
plot(m0_sc)
qqnorm(residuals(m0_sc))
qqline(residuals(m0_sc))  
dev.off()

library(DHARMa)
sim_res <- simulateResiduals(m0_sc)

pdf("plots/model_diagnostics_rr_simres.pdf", width = 8, height = 8)
plot(sim_res)    # QQ + residuals vs fitted
testOutliers(sim_res)
testDispersion(sim_res)                  
dev.off()


sink("tables/model_diagnostics_simres_m2.txt")
print(testOutliers(sim_res))
print(testDispersion(sim_res))
sink()


```

There are ~2× more outliers than expected under the model. This does not necessarily invalidate the model, but suggests the tails of the residual distribution are heavier than a Gaussian.
The GAM may handle this better, consider a response transformation (log) or a different family (e.g. Gamma if all values are positive).



This model is quite complex, and some predictors have confidence intervals that include zero, suggesting they may not be significant. 
The distribution of residuals is skewed, and there is a funnel shape that suggests heteroscedasticity.
It might be worth trying a GAM to allow for non-linear relationships, or at least removing some non-significant predictors to simplify the model.

A stepwise approach to model refinement can help identify a more parsimonious model that still explains the data well.

First, I will transform the response variable to see if that improves the residuals.


```{r transform_response}

wide_df <- wide_df %>%
  mutate(Recombination_rate_log = log10(Recombination_rate + 1e-6)) # add small constant to avoid log(0)

m0_log <- lmer(as.formula(paste("Recombination_rate_log ~", paste("break_status", "chr_status", paste(feat_cols_sc, collapse = " + "), sep = " + "), "+ (1 | genome)")),
           data = wide_df, REML = FALSE)


summary(m0_log)

pdf("plots/model_diagnostics_rr_log.pdf", width = 8, height = 8)
par(mfrow = c(1, 1))
residuals <- resid(m0_log)
hist(residuals, main = "Residuals Histogram", xlab = "Residuals", breaks = 30)
plot(m0_log)
qqnorm(residuals(m0_log))
qqline(residuals(m0_log))
dev.off()

anova(m0_sc, m0_log)

tidy(m0_sc, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Standardised effect on recombination rate", y = NULL) +
  theme_classic()

ggsave("plots/model_effects_rr_log.pdf", width = 8, height = 6)

```

With log transformation, the residuals look skewed in with a long tail in the negative direction, but the heteroscedasticity is reduced.
The qqnorm plot still shows strong deviation from normality, but but only at one end now. 
Kurtosis describes the shape of a distribution tails relative to a normal distribution, both tails curve away form the normal line in the qqplot, which suggests heavier tails than normal.



```{refine_model}
coef_tbl <- tidy(m0_sc, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)")

non_sig <- coef_tbl %>%
  filter(conf.low < 0 & conf.high > 0) %>%
  pull(term)

non_sig

feat_cols_sc_sig <- feat_cols_sc[!feat_cols_sc %in% non_sig]

m1_sc <- lmer(as.formula(paste("Recombination_rate ~",
                                paste(feat_cols_sc_sig, collapse = " + "), 
                                "+ (1 | genome)")),
              data = wide_df, REML = FALSE)

summary(m1_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=m1_sc =\n")
print(summary(m1_sc))
sink()

```

```{r}
# Compare models
anova(m0_sc, m1_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=Compare models=\n")
print(anova(m0_sc, m1_sc))
sink()

```

AIC and BIC suggest that the simpler model (m1_sc) is preferred, but the p-value is borderline.


```{r}
# Try adding back break_status nd chr_status, they are both significant in the full model and biologically relevant, but they are also related to each other, so I will add them back together.
m2_sc <- update(m1_sc, . ~ . + break_status + chr_status)
anova(m1_sc, m2_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=Compare models: Added break_status and chr_status=\n")
print(anova(m1_sc, m2_sc))
sink()

```


This significantly inproves the model fit. 



```{r}
# Try adding back Genes_z 
m3_sc <- update(m2_sc, . ~ . + Genes_z)
anova(m2_sc, m3_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=Compare models: Added Genes_z=\n")
print(anova(m2_sc, m3_sc))
sink()

```

Very limited effect on the model, and not significant, so I do not add Genes_z back in.

```{r}

m4_sc <- update(m2_sc, . ~ . + rel_position_z)
anova(m2_sc, m4_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=Compare models: Added rel_position_z=\n")
print(anova(m2_sc, m4_sc))
sink()



```

Relative position does not seem to have a significant effect when added back to the model with break_status, 
so I would not include it.


```{r}

m5_sc <- update(m2_sc, . ~ . + chr_length_z)
anova(m2_sc, m5_sc)

sink("tables/model_summaries.txt", append = TRUE)
cat("\n=Compare models: Added chr_length_z=\n")
print(anova(m2_sc, m5_sc))
sink()

```

Adding chromosome length back in has a borderline significant effect on the model, but it is not significant at the 0.05 level, so I do not include it either.


```{r diagnostics}

pdf("plots/model_diagnostics_rr_m2.pdf", width = 8, height = 8)
par(mfrow = c(1, 1))

residuals <- resid(m2_sc)
hist(residuals, main = "Residuals Histogram", xlab = "Residuals", breaks = 30)
plot(m2_sc)
qqnorm(residuals(m2_sc))
qqline(residuals(m2_sc))  
dev.off()

library(DHARMa)
sim_res <- simulateResiduals(m2_sc)

pdf("plots/model_diagnostics_rr_simres_m2.pdf", width = 8, height = 8)
plot(sim_res)                      # QQ + residuals vs fitted
testOutliers(sim_res)
testDispersion(sim_res)
dev.off()

```


## Non-linear model

Since many of the relationships between predictors and recombination rate are likely non-linear, a GAM can be a better choice to model these relationships.
Generalised Additive Model uses smooth functions of predictors instead of the predictors themselves.

gam or bam() in mgcv (nlme::gam is not recommended for large datasets)


s() defines a smooth term — it lets the relationship between a predictor and the response be non-linear, fitted as a spline.


How flexible the curve is is controlled automatically by the penalty (REML/GCV) — it finds the smoothest curve that still fits the data, preventing overfitting. 
The edf (effective degrees of freedom) in summary() tells you how non-linear the fitted curve is:

edf	Meaning
~1	Essentially linear
~2	Quadratic-like
>3	Clearly non-linear


```{r non_linar_model_gam0}

library(mgcv)
library(marginaleffects)

wide_df$genome <- as.factor(wide_df$genome)
wide_df$chr_category <- as.factor(wide_df$chr_category)

smooth_terms <- paste0("s(", feat_cols_sc, ")", collapse = " + ")

gam0 <- gam(
  as.formula(paste("Recombination_rate ~ s(genome, bs = 're') + break_status  +", smooth_terms)),
  data   = wide_df,
  method = "REML"           # fast REML, recommended for large data
)

summary(gam0)

sink(paste("tables/non_linear_model_summaries_", Sys.Date(), ".txt", sep = ""))
cat("\n=GAM=\n")
print(summary(gam0))
sink()

gam.check(gam0)

sink(paste("tables/gam_diagnostics_rr_", Sys.Date(), ".txt", sep = ""))
gam.check(gam0)
sink()

pdf(paste("plots/gam_diagnostics_rr2_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
plot.gam(gam0, pages = 1, shade = TRUE)
dev.off()

pdf(paste("plots/gam_marginal_effects_rr_rel_pos_", Sys.Date(), ".pdf", sep = ""),
    width = 8, height = 8)
plot_predictions(gam0, condition = c("rel_position_z", "genome", "chr_status"), 
                type = "response") +
  labs(x = "Relative position (z-score)", y = "Recombination rate",
                title = "Marginal effects of rel_position_z on recombination rate")
dev.off()

pdf("plots/gam_marginal_effects_rr_chr_length.pdf", width = 8, height = 8)
plot_predictions(gam0, condition = c("chr_length_z", "genome", "chr_status"), 
                type = "response") +
  labs(x = "Chromosome length (z-score)", y = "Recombination rate",
                title = "Marginal effects of chr_length_z on recombination rate") 
dev.off()

```


```{r gam_gamma_distribution}

gam0_gamma <- gam(
  as.formula(paste("Recombination_rate ~ s(genome, bs = 're') + break_status + chr_category.x +", smooth_terms)),
  data   = wide_df,
  method = "REML",           # fast REML, recommended for large data
  family = Gamma(link = "log")
)

summary(gam0_gamma)

sink(paste("tables/non_linear_model_summaries_", Sys.Date(), ".txt",sep = ""), append = TRUE)
cat("\n=GAM Gamma=\n")
print(summary(gam0_gamma))
print(gam.check(gam0_gamma))
sink()


pdf(paste("plots/gam_diagnostics_rr_gamma_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
plot.gam(gam0_gamma, pages = 1, shade = TRUE)
dev.off()

#Check residuals visually
pdf(paste("plots/gam_diagnostics_rr_gamma_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
par(mfrow = c(2, 2))
gam.check(gam0_gamma)
dev.off()


AIC(gam0, gam0_gamma)
BIC(gam0, gam0_gamma)

sink(paste("tables/non_linear_model_summaries_", Sys.Date(), ".txt",sep = ""), append = TRUE)
cat("\n=Model comparison=\n")
print(AIC(gam0, gam0_gamma))
print(BIC(gam0, gam0_gamma))
sink()


```

Both AIC and BIC suggest that the GAM with Gamma family is preferred over the GAM with Gaussian family, so I will proceed with the Gamma GAM for further refinement.


```{r refine_gam_gamma}

# only significant predictors

gam1 <- gam(
  Recombination_rate ~ 
    break_status + chr_category.x +
    s(genome, bs = "re") + s(chr_length_z) + s(dist_telo_z) + s(dist_bp_kb_z) +
    s(GC_z) + s(Genes_z) + s(LINE_z) + s(LTR_z) + s(Overlapping_z) + s(RC_z) + s(SINE_z) +
    s(Unknown_z) + s(eig_scores_z),
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam1)


AIC(gam0_gamma, gam1)
BIC(gam0_gamma, gam1)

sink(paste("tables/non_linear_model_summaries_", Sys.Date(), ".txt",sep = ""), append = TRUE)
cat("\n=GAM Gamma reduced=\n")
print(summary(gam1))
print(gam.check(gam1))
cat("\n=Model comparison GAM Gamma full vs reduced=\n")
print(AIC(gam0_gamma, gam1))
print(BIC(gam0_gamma, gam1))
sink()


pdf(paste("plots/gam_diagnostics_rr_gamma_reduced_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
plot.gam(gam1, pages = 1, shade = TRUE)
dev.off()

#Check residuals visually
pdf(paste("plots/gam_diagnostics_rr_gamma_reduced2_", Sys.Date(), ".pdf", sep = ""), width = 8, height = 8)
par(mfrow = c(2, 2))
gam.check(gam1)
dev.off()

# add variable distance to breakpoint



```

The reduced model has slightly higher AIC but strongly reduced BIC, so I would prefer the reduced model for parsimony.
The significance of the smooth terms is mostly unchanged, but the model is simpler and easier to interpret.

```{r}

gam1_mod <- gam(
  Recombination_rate ~ 
    break_status +
    s(genome, bs = "re") + s(chr_length_z) + s(dist_telo_z) + s(dist_bp_kb_z) +
    GC_z + Genes_z + s(LINE_z) + s(LTR_z) + s(Overlapping_z) + s(RC_z) + s(SINE_z) +
    s(Unknown_z) + s(eig_scores_z) + s(E3_z),
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam1_mod)

gam.check(gam1_mod)

print(AIC(gam1, gam1_mod))
print(BIC(gam1, gam1_mod))

```

```{r}
gam1_mod_rpt <- gam(
  Recombination_rate ~ 
    break_status +
    s(genome, bs = "re") + s(chr_length_z) + s(dist_telo_z) + s(dist_bp_kb_z) +
    GC_z + Genes_z + s(Total_repeats) + s(eig_scores_z) + s(E3_z),
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam1_mod_rpt)

gam.check(gam1_mod_rpt)

print(AIC(gam1, gam1_mod_rpt))
print(BIC(gam1, gam1_mod_rpt))
```



And including chr size in doubtfull beacuse the expected relationship between chromosome size and recombination rate is not well captured by the smooth term for chromosome size in the GAM, 
suggesting it is not a significant predictor of recombination rate in this case.

summary(gam_chr_f)

Family: Gamma 
Link function: log 

Formula:
Recombination_rate ~ break_status + chr_status + s(genome, bs = "re") + 
    s(chr_f, bs = "re") + s(GC_z) + s(LINE_z) + s(LTR_z) + s(Overlapping_z) + 
    s(Unknown_z) + s(eig_scores_z) + s(dist_telomere_z) + s(chr_length_z)

Parametric coefficients:
                           Estimate Std. Error t value Pr(>|t|)    
(Intercept)                -1.10443    0.11649  -9.481  < 2e-16 ***
break_statusfission        -0.04344    0.20029  -0.217 0.828321    
break_statusfission;fusion -0.15324    0.13026  -1.176 0.239484    
break_statusfusion         -0.30230    0.08129  -3.719 0.000202 ***
chr_statusrearranged       -0.10318    0.05107  -2.020 0.043407 *  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Approximate significance of smooth terms:
                       edf Ref.df        F  p-value    
s(genome)           0.9539  1.000 1630.694 6.40e-07 ***
s(chr_f)           20.3597 24.000   14.254  < 2e-16 ***
s(GC_z)             1.0008  1.002   10.663  0.00109 ** 
s(LINE_z)           4.6500  5.738   46.825  < 2e-16 ***
s(LTR_z)            1.0007  1.001   23.351 1.64e-06 ***
s(Overlapping_z)    1.0052  1.010   27.325 4.20e-07 ***
s(Unknown_z)        4.4979  5.494    6.287 3.44e-06 ***
s(eig_scores_z)     5.7421  6.965   28.143  < 2e-16 ***
s(dist_telomere_z)  8.7815  8.986   82.140  < 2e-16 ***
s(chr_length_z)     2.1464  2.192    1.448  0.18766    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

R-sq.(adj) =  0.302   Deviance explained = 32.5%
-REML = -2928.9  Scale est. = 0.18237   n = 4983




```{r plot_position_vs_rec_rate}
ggarrange(
ggplot(wide_df, aes(x = dist_telomere, y = Recombination_rate)) +
  geom_point(alpha = 0.1) +
  geom_smooth(aes(colour = chr), method = "loess", se = FALSE) +
  facet_wrap(~ genome) +
  labs(x = "Distance to telomere", y = "Recombination rate") +
  theme_classic(),

ggplot(wide_df, aes(x = rel_position, y = Recombination_rate)) +
  geom_point(alpha = 0.1) +
  geom_smooth(aes(colour = chr), method = "loess", se = FALSE) +
  facet_wrap(~ genome) +
  labs(x = "Relative position", y = "Recombination rate") +
  theme_classic(),
ncol = 1, align = "v", common.legend = TRUE, legend = "bottom"
)
ggsave("plots/rel_pos_dist_telo_vs_rec_rate.pdf", width = 12, height = 16)


#Check if median rec rate per chromosome is correlate with chromosome lenght and if the chromosome status is relevant for this correlation

pdf("plots/rec_rate_chr_length.pdf", width = 8, height = 6)
ggplot(wide_df %>% group_by(genome, chr) %>% summarise(median_rec_rate = median(Recombination_rate, na.rm = TRUE),
                                                        chr_length = first(chr_length),
                                                        chr_status = first(chr_status)),
       aes(x = chr_length, y = median_rec_rate, color = chr_status)) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
  labs(x = "Chromosome length", y = "Median recombination rate") +
  theme_classic()

dev.off()
```


```{r rel_position_effects}


pdf("plots/gam_marginal_effects_rr_rel_pos_reduced.pdf", width = 8, height = 8)
plot_predictions(gam1, variables = "rel_position_z", 
  condition = c("rel_position_z", "genome", "chr_status"), 
  type = "response") +
  labs(x = "Relative position, scaled", y = "Effect on recombination rate",
       title = "Marginal effects of rel_position_z on recombination rate")
dev.off()

# plot derivative of the smooth term for rel_position_z

pdf("plots/gam_marginal_effects_rr_rel_pos_reduced_derivative.pdf", width = 8, height = 8)
plot_slopes(gam1, variables = "rel_position_z", 
  condition = c("rel_position_z", "genome", "chr_status"), 
  type = "response") + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Relative position, scaled", y = "1st derivative of effect on recombination rate",
       title = "Marginal effects of rel_position_z on recombination rate")
dev.off()



########## Not run

# Derivative of the smooth for rel_position_z
deriv_rel <- derivatives(gam1, term = "s(rel_position_z)", n = 500)


# Where derivative confidence interval includes 0 = slope not different from flat
asymptote_region <- deriv_rel %>%
  filter(.lower_ci < 0 & .upper_ci > 0)   # CI includes zero = plateau

cat("Asymptote region (z-score):", range(asymptote_region$rel_position_z), "\n")

# Back-transform to original scale if needed
rel_pos_mean <- mean(wide_df$rel_position, na.rm = TRUE)
rel_pos_sd   <- sd(wide_df$rel_position, na.rm = TRUE)
cat("Asymptote region (original scale):", 
    range(asymptote_region$rel_position_z) * rel_pos_sd + rel_pos_mean, "\n")

pred_rel <- predictions(gam1, newdata = datagrid(rel_position_z = seq(min(wide_df$rel_position_z, na.rm=TRUE), 
                                                                        max(wide_df$rel_position_z, na.rm=TRUE), 
                                                                        length.out = 200)))

p1 <- ggplot(pred_rel, aes(x = rel_position_z, y = estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "rel_position_z", y = "Recombination rate") +
  theme_classic()

p2 <- ggplot(deriv_rel, aes(x = rel_position_z, y = .derivative)) +
  geom_line() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "rel_position_z", y = "Slope") +
  theme_classic()


p1 / p2
ggsave("plots/rel_position_asymptote.pdf", width = 8, height = 8)

```


```{r dist_telomere_effects}
# replace rel_position with dist_telomere in the GAM to see if the distance to telomere has a non-linear effect on recombination rate, and if there is an asymptote

wide_df$chr_f <- factor(paste(wide_df$genome, wide_df$chr, sep = "_"))

gam_dist_tel <- gam(
  Recombination_rate ~ 
    break_status + chr_status +
    s(genome, bs = "re") +
    s(GC_z) + s(LINE_z) + s(LTR_z) + s(Overlapping_z) + 
    s(Unknown_z) + s(eig_scores_z) + s(dist_telomere_z) +
    s(chr_length_z),
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam_dist_tel)
gam.check(gam_dist_tel)

gam_dist_tel_lin <- gam(
  Recombination_rate ~ 
    break_status + chr_status +
    s(genome, bs = "re") +
    s(GC_z, k=1) + s(LINE_z) + s(LTR_z, k=1) + s(Overlapping_z) + 
    s(Unknown_z) + s(eig_scores_z) + s(dist_telomere_z) +
    s(chr_length_z),
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam_dist_tel_lin)
gam.check(gam_dist_tel_lin)



```


```{r dist_telomere_effects_asymtote}
# Derivative of the smooth for dist_telomere_z
deriv_mb <- derivatives(gam_dist_tel, term = "s(dist_telomere_z)", n = 500)
deriv_rel <- derivatives(gam1, term = "s(rel_position_z)", n = 500)


# Where derivative confidence interval includes 0 = slope not different from flat
asymptote_region_mb <- deriv_mb %>% filter(.lower_ci < 0 & .upper_ci > 0)   # CI includes zero = plateau
asymptote_region_rel <- deriv_rel %>% filter(.lower_ci < 0 & .upper_ci > 0)   # CI includes zero = plateau

cat("Asymptote region (z-score):", range(asymptote_region_mb$dist_telomere_z), "\n")
cat("Asymptote region (z-score):", range(asymptote_region_rel$rel_position_z), "\n")

# Back-transform to original scale if needed
dist_telomere_mean <- mean(wide_df$dist_telomere, na.rm = TRUE)
dist_telomere_sd   <- sd(wide_df$dist_telomere, na.rm = TRUE)
cat("Asymptote region (original scale):", 
    range(asymptote_region$dist_telomere_z) * dist_telomere_sd + dist_telomere_mean, "\n")

pred_rel <- predictions(gam_dist_tel, newdata = datagrid(dist_telomere_z = seq(min(wide_df$dist_telomere_z, na.rm=TRUE), 
                                                                        max(wide_df$dist_telomere_z, na.rm=TRUE), 
                                                                        length.out = 200)))


p1 <- ggplot(pred_rel, aes(x = dist_telomere_z, y = estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "dist_telomere_z", y = "Recombination rate") +
  theme_classic()

p2 <- ggplot(deriv_rel, aes(x = dist_telomere_z, y = .derivative)) +
  geom_line() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "dist_telomere_z", y = "Slope") +
  theme_classic()


p1 / p2
ggsave("plots/dist_telomere_asymptote.pdf", width = 8, height = 8)
```


```{r dist_telomere_effects_asymtote_chr}

wide_df$chr_f <- factor(paste(wide_df$genome, wide_df$chr, sep = "_"))

gam_telo_chr <- gam(
  Recombination_rate ~
    break_status + s(genome, bs = "re") +
    s(GC_z) + s(LINE_z) + s(LTR_z) + s(Overlapping_z) +
    s(Unknown_z) + s(eig_scores_z) + s(chr_length_z) +
    s(dist_telomere_z, chr_f, bs = "fs"),  # random smooths per chromosome
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam_telo_chr)

telo_mean <- mean(wide_df$dist_telomere, na.rm = TRUE)
telo_sd   <- sd(wide_df$dist_telomere, na.rm = TRUE)

# Get smooth estimates per chromosome from the factor smooth

sm_est <- smooth_estimates(gam_telo_chr,
                           smooth = "s(dist_telomere_z,chr_f)",
                           n = 200) %>%
  add_confint()

# Numerical derivative per chromosome
deriv_chr <- sm_est %>%
  group_by(chr_f) %>%
  arrange(dist_telomere_z) %>%
  mutate(
    d_dx            = c(diff(.estimate) / diff(dist_telomere_z), NA),
    dist_telomere_bp = dist_telomere_z * telo_sd + telo_mean
  ) %>%
  filter(!is.na(d_dx))

# Asymptote: first position where slope magnitude drops into the bottom 20%
asymptotes_chr <- deriv_chr %>%
  group_by(chr_f) %>%
  mutate(flat_threshold = quantile(abs(d_dx), 0.2)) %>%
  filter(abs(d_dx) < flat_threshold) %>%
  summarise(
    asym_start_bp = min(dist_telomere_bp),
    asym_start_z  = min(dist_telomere_z),
    .groups = "drop"
  )

print(asymptotes_chr)

# Add genome info for species comparison (chr_f levels are per genome)
asymptotes_chr <- asymptotes_chr %>%
  left_join(wide_df %>% distinct(chr_f, genome, chr, chr_length), by = "chr_f")


# Plot smooth curve per chromosome, with asymptote marked
ggplot(sm_est %>% mutate(dist_telomere_bp = dist_telomere_z * telo_sd + telo_mean),
       aes(x = dist_telomere_bp, y = .estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_vline(data = asymptotes_chr,
             aes(xintercept = asym_start_bp), linetype = "dotted", colour = "blue") +
  facet_wrap(~ chr_f, scales = "free_y") +
  labs(x = "Distance to telomere (bp)", y = "Smooth effect on recombination rate") +
  theme_classic()
ggsave("plots/dist_telomere_smooth_per_chr.pdf", width = 16, height = 12)

# plot the asymtote start as a function of chr size, to see if larger chromosomes have a longer region of increasing recombination rate from the telomere
# plot the distance distribution of the asymptote start for all chromosomes, and see if there is a common distance from the telomere where the recombination rate starts to plateau
# add text at median asymptote start distance to show the typical distance from the telomere where recombination rate starts to plateau


ggarrange(
      {
      asym_filt <- asymptotes_chr %>% filter(asym_start_bp < 4e6)
      med_asym  <- median(asymptotes_chr$asym_start_bp / 1e6)
      ggplot(asym_filt, aes(x = asym_start_bp / 1e6)) +
        geom_histogram(bins = 15, fill = "steelblue", colour = "black") +
        geom_vline(xintercept = med_asym, linetype = "dashed", colour = "black") +
        annotate("text", x = med_asym, y = Inf,
                 label = paste0("Median: ", round(med_asym, 4), " Mb"),
                 vjust = 2, hjust = -0.8, colour = "black") +
        labs(x = "Asymptote start (Mb from telomere), outlier removed", y = "Count") +
        theme_classic()
    },

    ggplot(asymptotes_chr, aes(x = chr_length/1e6, y = asym_start_bp/1e6)) +
      geom_point(aes(colour = genome)) +
      geom_smooth(method = "lm") +
      stat_cor(method = "spearman", label.x.npc = "centre", label.y.npc = "top") +
      labs(x = "Chromosome length (Mb)", y = "Asymptote start (Mb from telomere)") +
      theme_classic(),

    ncol = 2, align = "h"
) 

ggsave("plots/asymptote_start_distr_vs_chr_length.pdf", width = 10, height = 4)

```


```{r rel_pos_effects_asymtote_chr}

wide_df$chr_f <- factor(paste(wide_df$genome, wide_df$chr, sep = "_"))

gam_telo_chr <- gam(
  Recombination_rate ~
    break_status + s(genome, bs = "re") +
    s(GC_z) + s(LINE_z) + s(LTR_z) + s(Overlapping_z) +
    s(Unknown_z) + s(eig_scores_z) + s(chr_length_z) +
    s(dist_telomere_z, chr_f, bs = "fs"),  # random smooths per chromosome
  data   = wide_df,
  method = "REML",
  family = Gamma(link = "log")
)

summary(gam_telo_chr)

telo_mean <- mean(wide_df$dist_telomere, na.rm = TRUE)
telo_sd   <- sd(wide_df$dist_telomere, na.rm = TRUE)

# Get smooth estimates per chromosome from the factor smooth

sm_est <- smooth_estimates(gam_telo_chr,
                           smooth = "s(dist_telomere_z,chr_f)",
                           n = 200) %>%
  add_confint()

# Numerical derivative per chromosome
deriv_chr <- sm_est %>%
  group_by(chr_f) %>%
  arrange(dist_telomere_z) %>%
  mutate(
    d_dx            = c(diff(.estimate) / diff(dist_telomere_z), NA),
    dist_telomere_bp = dist_telomere_z * telo_sd + telo_mean
  ) %>%
  filter(!is.na(d_dx))

# Asymptote: first position where slope magnitude drops into the bottom 20%
asymptotes_chr <- deriv_chr %>%
  group_by(chr_f) %>%
  mutate(flat_threshold = quantile(abs(d_dx), 0.2)) %>%
  filter(abs(d_dx) < flat_threshold) %>%
  summarise(
    asym_start_bp = min(dist_telomere_bp),
    asym_start_z  = min(dist_telomere_z),
    .groups = "drop"
  )

print(asymptotes_chr)

# Add genome info for species comparison (chr_f levels are per genome)
asymptotes_chr <- asymptotes_chr %>%
  left_join(wide_df %>% distinct(chr_f, genome, chr, chr_length), by = "chr_f")


# Plot smooth curve per chromosome, with asymptote marked
ggplot(sm_est %>% mutate(dist_telomere_bp = dist_telomere_z * telo_sd + telo_mean),
       aes(x = dist_telomere_bp, y = .estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_vline(data = asymptotes_chr,
             aes(xintercept = asym_start_bp), linetype = "dotted", colour = "blue") +
  facet_wrap(~ chr_f, scales = "free_y") +
  labs(x = "Distance to telomere (bp)", y = "Smooth effect on recombination rate") +
  theme_classic()
ggsave("plots/dist_telomere_smooth_per_chr.pdf", width = 16, height = 12)

# plot the asymtote start as a function of chr size, to see if larger chromosomes have a longer region of increasing recombination rate from the telomere
# plot the distance distribution of the asymptote start for all chromosomes, and see if there is a common distance from the telomere where the recombination rate starts to plateau
# add text at median asymptote start distance to show the typical distance from the telomere where recombination rate starts to plateau


ggarrange(
      {
      asym_filt <- asymptotes_chr %>% filter(asym_start_bp < 4e6)
      med_asym  <- median(asymptotes_chr$asym_start_bp / 1e6)
      ggplot(asym_filt, aes(x = asym_start_bp / 1e6)) +
        geom_histogram(bins = 15, fill = "steelblue", colour = "black") +
        geom_vline(xintercept = med_asym, linetype = "dashed", colour = "black") +
        annotate("text", x = med_asym, y = Inf,
                 label = paste0("Median: ", round(med_asym, 4), " Mb"),
                 vjust = 2, hjust = -0.8, colour = "black") +
        labs(x = "Asymptote start (Mb from telomere), outlier removed", y = "Count") +
        theme_classic()
    },

    ggplot(asymptotes_chr, aes(x = chr_length/1e6, y = asym_start_bp/1e6)) +
      geom_point(aes(colour = genome)) +
      geom_smooth(method = "lm") +
      stat_cor(method = "spearman", label.x.npc = "centre", label.y.npc = "top") +
      labs(x = "Chromosome length (Mb)", y = "Asymptote start (Mb from telomere)") +
      theme_classic(),

    ncol = 2, align = "h"
) 

ggsave("plots/asymptote_start_distr_vs_chr_length.pdf", width = 10, height = 4)

```


```{r plot_ms}

rec_com_df  <- combined_df_brp %>% filter(feature == "Recombination_rate" & !(chr %in% c("Z", "Z1", "Z2", "W", "W1", "W2"))) 

textsize=12
chr_ends <- 
rec_com_df %>% filter(!is.na(chr_category)) %>%
  ggplot(aes(rel_position, value)) +
  #geom_point() +
  geom_smooth(aes(colour = genome, linetype = chr_category), se = F) +
    scale_x_continuous(name = "Relative distance from telomere") +
  scale_y_continuous(name = "Recomb. rate (cM/Mb)",
                     limits = c(0,6)) +
  scale_linetype_manual(values = c("solid", "dashed"),
                        labels = c("Collinear", "Rearranged"),
                        name = "Chromosome category") +
  scale_colour_manual(name = "Species",
                      values = mec_colors,
                      labels = c(expression(italic("M. lysimnia")), expression(italic("M. polymnia"))),
                      guide = "none") +
    theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 0.5),
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size=textsize),
        legend.title = element_text(size=textsize),
        axis.text = element_text(size=textsize),
        axis.title = element_text(size=textsize),
        )

rec_brp <- 
rec_com_df %>% filter(!closest_break_type %in% c("colinear", "shared_fission")) %>%
  ggplot(aes(dist_bp_kb/1000000, value)) +
  #geom_point() +
  geom_smooth(aes(linetype = closest_break_type, colour = genome), se = F, linewidth = 1) +
  scale_x_continuous(limits = c(0, 5),
                     name = "Distance from rearrangement breakpoint (Mb)") +
  scale_y_continuous(name = "Recomb. rate (cM/Mb)",
                     limits = c(0,6)) +
  scale_linetype_manual(values = c( "dotted","dashed", "solid"),
                        labels = c("Fission", "Fission-fusion", "Fusion"),
                        name = "Breakpoint") +
  scale_colour_manual(name = "Species",
                      values = mec_colors,
                      labels = c(expression(italic("M. lysimnia")), expression(italic("M. polymnia"))),
                      guide = "none") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 0.5),
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size=textsize),
        legend.title = element_text(size=textsize),
        axis.text = element_text(size=textsize),
        axis.title = element_text(size=textsize),
        )

View(rec_com_df)

fill_colors <- c(
  "colinear.ilMecPoly1"   = "#2166AC", # Custom species color
  "rearranged.ilMecPoly1" = "white",
  "colinear.ilMecLysi212"   = "darkorange", # Custom species color
  "rearranged.ilMecLysi212" = "white"
)

chr_med_rr <- rec_com_df %>% filter(feature == "Recombination_rate") %>% group_by(genome, chr, chr_length) %>% summarise(med_rec_rate=median(value)) %>% left_join(., chr_class) %>% filter(!is.na(chr_category))

label_sp <- labeller(genome = c(ilMecPoly1 ="M. polymnia",
                                   ilMecLysi212 = "M. lysimnia"))

chr_rr_box <- chr_med_rr %>%
    ggplot() +
    geom_boxplot(aes(chr_category, med_rec_rate, fill = interaction(chr_category, genome), colour = genome), alpha = 0.6, outliers = F) +
  geom_jitter(aes(chr_category, med_rec_rate, fill = interaction(chr_category, genome), colour = genome)) +
  facet_wrap(~genome, strip.position = "bottom", labeller = label_sp) +  
  scale_y_continuous(name = "Recomb. rate (cM/Mb)",
                     limits = c(2,6)) +
  scale_fill_manual(
    name = "Chromosome category",
    values = fill_colors,
    breaks = c("colinear.ilMecPoly1", "rearranged.ilMecPoly1"),
    labels = c("Collinear","Rearranged")
    ) +
  scale_colour_manual(name = "Species",
                      values = mec_colors,
                      labels = c(expression(italic("M. lysimnia")), expression(italic("M. polymnia")))) +
   annotate(geom = "text", label = "Adj.R[2] = 0.305\np < 0.001", 
            #x = 5.5, y = 33, 
           color = "black", 
           #hjust = 0, 
           size = 4) +
  #coord_cartesian(clip = "off") +
    theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 0.5),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text = element_text(size=textsize, face = "italic"),
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size=textsize),
        legend.title = element_text(size=textsize),
        axis.text.y = element_text(size=textsize),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size=textsize),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank()
        )


label_cat <- labeller(chr_category = c(colinear ="Collinear",
                                   rearranged = "Rearranged"))

chr_rr_point <- 
  chr_med_rr %>%
    ggplot(aes(chr_length/1000000, med_rec_rate)) +
    geom_point(aes(fill = interaction(chr_category, genome), colour = genome), shape = 21, size = 6, stroke = 1) +
  facet_wrap(~chr_category, labeller = label_cat) +  
  scale_y_continuous(name = "Recomb. rate (cM/Mb)",
                     limits = c(2,6)) +
  scale_fill_manual(
    name = "Chromosome category",
    values = fill_colors,
    breaks = c("colinear.ilMecPoly1", "rearranged.ilMecPoly1"),
    labels = c("Collinear","Rearranged")
    ) +
  scale_x_continuous(name = "Chromosome length (Mb)") +
  scale_colour_manual(name = "Species",
                      values = mec_colors,
                      labels = c(expression(italic("M. lysimnia")), expression(italic("M. polymnia")))) +
  guides(colour = guide_legend(
    override.aes = list(
      shape = 16        
    )
  )) +
    ggnewscale::new_scale_colour() +
    geom_text(aes(label = chr, colour = chr_category), show.legend = F) +
   geom_smooth(method = "lm", colour = "black", se = F) +
    scale_colour_manual(values = c("colinear" = "white", "rearranged" = "black")) +
    theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 0.5),
        strip.background = element_rect(fill = "white", colour = "black"),
        strip.text = element_text(size=textsize),
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size=textsize),
        legend.title = element_text(size=textsize),
        axis.text.y = element_text(size=textsize),
        axis.text.x = element_text(size=textsize),
        axis.title.y = element_text(size=textsize),
        axis.title.x = element_text(size=textsize),
        guides()) 




rec_fig3 <- 
ggarrange(
  ggarrange(chr_ends + theme(legend.position = "inside",
                             legend.position.inside = c(0.7, 0.2)), 
          rec_brp  + theme(legend.position = "inside",
                             legend.position.inside = c(0.7, 0.23)),
          labels = c("A", "B")),
  ggarrange(chr_rr_point, 
          chr_rr_box + theme(plot.margin = margin(20, 5, 11, 5)), 
          common.legend = T, legend = "top",
          labels = c("C", "D")),
  nrow = 2,
  heights = c(1,1.2)
)


  ggarrange(ggarrange(feat_div_box, (end_plot + theme(plot.margin = margin(30, 5, 0, 5))),
                                     nrow = 1, widths = c(1.5,1)),
          ggarrange(rec_comp_plot + theme(plot.margin = margin(5, 5, 10, 5)), (brp_comp_plot + theme(plot.margin = margin(5, 5, 0, 5))), nrow = 1, widths = c(1.5,1)),
          nrow = 2,
          heights = c(1.8,1),
          align = "v"
          )
  
  ggarrange(
    ggarrange(rec_fig3, 
              ggarrange(rec_comp_plot + theme(plot.margin = margin(5, 5, 10, 5)), (brp_comp_plot + theme(plot.margin = margin(5, 5, 0, 5))), nrow = 1, widths = c(1,1),
              labels = c("E","F")),
              nrow = 2, heights = c(2,1)),
    ggarrange(feat_div_box, (end_plot + theme(plot.margin = margin(30, 5, 0, 5))),
              nrow = 1, widths = c(1,1), 
              labels = c("G", "H")))

  ggsave(filename = paste("plots/fig_3_", Sys.Date(), ".pdf", sep = ""),
         height = 10,
         width = 18)  

  ggsave(filename = paste("plots/fig_3_", Sys.Date(), ".png", sep = ""),
         height = 10,
         width = 18)  
```

```{r}

chr_med_rr

summary(lm(med_rec_rate ~ chr_length*genome + chr_category*genome + chr_length*chr_category*genome, data = chr_med_rr))

model_full <- lm(med_rec_rate ~ chr_length * genome * chr_category, data = chr_med_rr)
summary(model_full)

summary(lm(med_rec_rate ~ chr_length * chr_category, data = chr_med_rr %>% filter(genome == "ilMecPoly1")))

summary(lm(med_rec_rate ~ chr_length * chr_category, data = chr_med_rr %>% filter(genome == "ilMecLysi212")))

chr_med_rr$chr_length_mb <- chr_med_rr$chr_length/1000000

mixed_model <- lmer(
  med_rec_rate ~ chr_length_mb * chr_category + (1 + chr_length_mb | genome), 
  data = chr_med_rr
)

library(lmerTest)
library(performance)
summary(simplified_mixed_model)
# Use this simplified version only if the first model throws a warning
simplified_mixed_model <- lmer(med_rec_rate ~ chr_length_mb * chr_category + (1 | genome), data = chr_med_rr)

model_performance(simplified_mixed_model)

```

