---
title: "vcfqc"
author: "KNasvall"
date: "2026-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, echo=FALSE}
#install.packages("vcfR")
library(vcfR)
library(tidyverse)
```

## Quality control of VCF files

Based on https://knausb.github.io/vcfR_documentation/quick_intro.html#Output_of_data

The vcf-files are thinned with bcftools using the following parameters:


```{r bash, echo=TRUE, eval=FALSE}
bcftools view -v snps ${file} | bcftools +prune -w 10000bp -n 1 -N rand --output ${OUT_FILE%.*}_thin_snps.vcf.gz -Oz
```

```{r getdata}
WD="/lustre/scratch125/tol/teams/meier/users/kn9/recombination/rawdata_to_vcf/04_vcf_filtering/polymnia/01_vcf_qc"

VCF_FILE="vcf_thinned/vcf_thin_all_snps.vcf.gz"

#FASTA="/lustre/scratch125/tol/teams/meier/users/kn9/recombination/rawdata_to_vcf/03_variant_call_lysimnia_rptmask/ref/ilMecLysi212_1.renamed.masked.fa.gz"

OUTPUT_DIR=WD
# crashed when trying to read in a small chr vcf file, use thinned files
#vcf_thin_pca.bsub # thinning with bcftools

# inspired from CAM0707355
col_pal_mechanitis <- c("orange3","#8A5628", "gold2", "#2D1A12","#807267", "#9E8757","#D6D5D6", "white")

text_size=20

```

```{r read_data, echo=FALSE}
vcf <- read.vcfR(VCF_FILE)
#vcf

# only retain non indels, not needed the thinning only include snps
# vcf2 <- extract.indels(vcf)

# only biallelic
vcf <- vcf[is.biallelic(vcf),]
vcf

# only needed if want to use chromosome functions
# dna <- ape::read.dna(FASTA, format = "fa")
# dna
```


```{r summary_vcf, echo=FALSE, eval=FALSE}
vcf
# just print a summary, underestimate missing data (all fields NA, so if something is recorded it will not count as missing even if ./.)
head(vcf) # summarize each section of the object

# vcf@meta # header content from the vcf file
# #summarisies the meta
# queryMETA(vcf)
# queryMETA(vcf, element = 'DP')
# 
# head(vcf@fix) # table with the first 8 columns of the vcf 
# head(getFIX(vcf)) #(the INFO col is suppressed, to make it more manageble)
# 
# vcf@gt[1:6, 1:4] # genotype and sample info, this showes first 6 variants of first 4 columns (FORMAT and the first 3 samples)
```

```{r funct_He_MAF_EHET, echo=FALSE} 
# Function computing minimum allele frequency per site, expected heterozygosity, and expected number of heterozygous sites per sample. The output are three named lists
#
compute_He_MAF_EHET <- function(gt) {
  # number of called genotypes per site
  n_called <- rowSums(!is.na(gt))

  # reference allele count
  ref_alleles <- rowSums(
    ifelse(gt == "0/0", 2,
           ifelse(gt == "0/1" | gt == "1/0", 1, 0)),
    na.rm = TRUE
  )
  
  # alternate allele count
  alt_alleles <- rowSums(
    ifelse(gt == "1/1", 2,
           ifelse(gt == "0/1" | gt == "1/0", 1, 0)),
    na.rm = TRUE
  )
  
  # allele frequencies
  p <- ref_alleles / (2 * n_called)
  q <- alt_alleles / (2 * n_called)
  
  # expected heterozygosity per site
  He <- 2 * p * q
  
  # minor allele frequency
  MAF <- pmin(p, q)
  
  # expected heterozygous sites per sample
  E_HET <- sapply(seq_len(ncol(gt)), function(j) {
    sum(He[!is.na(gt[, j])], na.rm = TRUE)
  })
  names(E_HET) <- colnames(gt)
  
  return(list(
    He = He,
    MAF = MAF,
    E_HET = E_HET
  ))
}




```



```{r chrom_object, eval=FALSE, echo=FALSE}
#Skipping the chrom object can only read one chr at the time
chrom <- create.chromR(name = 'test_lysimnia',
                       vcf = vcf, 
                       seq = dna)

```


```{r plot_chrom, eval=FALSE, echo=FALSE}
plot(chrom)

```

```{r chromoqc, eval=FALSE, echo=FALSE}
chromoqc(chrom)
```

```{r process_chrom, eval=FALSE, echo=FALSE}
chrom <- proc.chromR(chrom, verbose = TRUE, win.size = 100000)
chromoqc(chrom)
plot(chrom)
```


## Variant based statistics

Variant quality, depth, missingness and minimum allele frequency (MAF)

### Variant quality     

```{r variant_quality, echo=FALSE}

DP <- getFIX(vcf, getINFO = TRUE)
df <- as.data.frame(DP, stringsAsFactors = FALSE)

df2 <- df[, c("CHROM", "POS", "QUAL")]
df2$POS <- as.numeric(df2$POS)
df2$QUAL <- as.numeric(df2$QUAL)

df2$CHROM <- sub("SUPER_", "", df2$CHROM)
# zeropad
pad_numbers <- function(x, width = 2) {
  is_num <- grepl("^\\d+$", x)          # TRUE for only digits
  x[is_num] <- sprintf(paste0("%0", width, "d"), as.integer(x[is_num]))
  x
}

df2$CHROM <- pad_numbers(df2$CHROM, width = 2)


p <- df2 %>%
  ggplot() +
  geom_point(aes(POS/1000000, QUAL), colour = col_pal_mechanitis[2], alpha = 0.5, size = 0.5) +
  facet_wrap(~CHROM) +
  scale_x_continuous(name = "Position (Mb)") +
  theme_classic() +
  theme(panel.grid.major.y = element_line(colour = "grey"), 
        strip.background = element_rect(fill = col_pal_mechanitis[7]))

p

df2 %>%
  ggplot() +
  geom_histogram(aes(QUAL), colour = col_pal_mechanitis[4], fill = col_pal_mechanitis[2], alpha = 0.5) +
  facet_wrap(~CHROM) +
  #scale_x_continuous(limits = c(0,100)) +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        strip.background = element_rect(fill = col_pal_mechanitis[7]))

df2 %>%
  ggplot() +
  geom_histogram(aes(QUAL), colour = col_pal_mechanitis[4], fill = col_pal_mechanitis[2], alpha = 0.5) +
  facet_wrap(~CHROM) +
  scale_x_continuous(limits = c(0,100), name = "QUAL, capped to 100") +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        strip.background = element_rect(fill = col_pal_mechanitis[7]))



```



### Variant depth

```{r var_depth,  echo=FALSE}

df2$INFO_DP <- as.numeric(sub(".*DP=([0-9]+).*", "\\1", df$INFO))

df2 %>%
  ggplot() +
  geom_histogram(aes(INFO_DP), colour = col_pal_mechanitis[4], fill = col_pal_mechanitis[2], alpha = 0.5) +
  facet_wrap(~CHROM) +
  #scale_x_continuous(limits = c(0,2000)) +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank(),
        strip.background = element_rect(fill = col_pal_mechanitis[7]))

print("Summary INFO/DP:")
summary(df2$INFO_DP)

df2 %>%
  ggplot() +
  geom_histogram(aes(INFO_DP), colour = col_pal_mechanitis[4], fill = col_pal_mechanitis[2], alpha = 0.5) +
  facet_wrap(~CHROM) +
  scale_x_continuous(name = "INFO_DP, capped to 3*median",
                     limits = c(0,median(df2$INFO_DP)*3)) +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank(),
        strip.background = element_rect(fill = col_pal_mechanitis[7]))


df2 %>%
  ggplot() +
  geom_point(aes(POS/1000000, INFO_DP), colour = col_pal_mechanitis[2], alpha = 0.5) +
  facet_wrap(~CHROM) +
  scale_x_continuous(name = "Position (Mb)") +
  scale_y_continuous(limits = c(0,median(df2$INFO_DP)*3)) +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank(),
        strip.background = element_rect(fill = col_pal_mechanitis[7]))






```


### Variant missingness

https://knausb.github.io/vcfR_documentation/missing_data.html


```{r miss_hist_sites, echo=FALSE}
gt_mat <- extract.gt(vcf, element = "GT")

varMiss <- apply(gt_mat, MARGIN = 1, function(x){ sum(is.na(x)) })
varMiss <- varMiss/ncol(vcf@gt[,-1])

hist(varMiss, col = "gold3", xlab = "Missingness (%)")

```



### Variant MAF

```{r variant_MAF, echo=FALSE}

gt <- extract.gt(vcf, element = "GT")

He_MAF_EHET <- compute_He_MAF_EHET(gt)

summary(He_MAF_EHET$MAF)
hist(He_MAF_EHET$MAF, col = "gold3", xlab = "Minimum allele frequency", breaks = 1000, xlim = c(0,0.01))


print("Recommended MAF if at least present in two haplotypes in the sample: ")
2/(length(gt[1,])*2)

```



## Individual-based metrics

Genotype quality, depth, missingness (missing sites per individual), F (inbreeding coefficient)

### Genotype quality

```{r GQ, echo=FALSE}

gq <- extract.gt(vcf, element = "GQ", as.numeric = T)

# remove missing sites
gq[gt == "./." | is.na(gt)] <- NA


gq.df <- as.data.frame(gq)
gq.df$variant_id <- rownames(gq)
gq.df$CHROM <- sapply(str_split(gq.df$variant_id, "_"), `[`, 2)

# zeropad
pad_numbers <- function(x, width = 2) {
  is_num <- grepl("^\\d+$", x)          # TRUE for only digits
  x[is_num] <- sprintf(paste0("%0", width, "d"), as.integer(x[is_num]))
  x
}

gq.df$CHROM <- pad_numbers(gq.df$CHROM, width = 2)


gq.df$POS <- sapply(str_split(gq.df$variant_id, "_"), `[`, 3)
gq.df$POS <- as.numeric(gq.df$POS)


gq.df <- gq.df %>% pivot_longer(
    cols = -c(variant_id,CHROM,POS),
    names_to = "sample",
    values_to = "quality"
  ) %>% 
  group_by(sample) %>%
  mutate(med_gq = median(quality, na.rm = TRUE))

summary(gq.df$quality)

print("Samples with median GQ<30:")
unique(gq.df[gq.df$med_gq<30,"sample"])


hist(gq.df$quality, col = "#D6D5D6", xlab = "Genotype quality", breaks = 100)

gq.df %>%
  ggplot() +
  #geom_violin(aes(sample, depth)) +
  geom_boxplot(aes(reorder(sample, med_gq, decreasing = T), quality, fill = med_gq<30 ), outliers = FALSE) +
  #geom_point(aes(sample, depth), alpha = 0) +
  scale_y_continuous(#position = "right",
                     #limits = c(0,50),
                     name = "Quality") +
  scale_fill_manual(values = c(col_pal_mechanitis[7], "orange")) +
  coord_flip() +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank())


```



### Individual depth
```{r ind_depth, echo = FALSE}
dp <- extract.gt(vcf, element = "DP", as.numeric=TRUE)

# remove missing sites
#dp[gt == "./." | is.na(gt)] <- NA

dp.df <- as.data.frame(dp)
dp.df$variant_id <- rownames(dp)
dp.df$CHROM <- sapply(str_split(dp.df$variant_id, "_"), `[`, 2)

# zeropad
pad_numbers <- function(x, width = 2) {
  is_num <- grepl("^\\d+$", x)          # TRUE for only digits
  x[is_num] <- sprintf(paste0("%0", width, "d"), as.integer(x[is_num]))
  x
}

dp.df$CHROM <- pad_numbers(dp.df$CHROM, width = 2)


dp.df$POS <- sapply(str_split(dp.df$variant_id, "_"), `[`, 3)
dp.df$POS <- as.numeric(dp.df$POS)


dp.df <- dp.df %>% pivot_longer(
    cols = -c(variant_id,CHROM,POS),
    names_to = "sample",
    values_to = "depth"
  ) %>% 
  group_by(sample) %>%
  mutate(med_dp = median(depth, na.rm = TRUE))

hist(dp.df$depth, col = "#807267", xlab = "Individual depth", breaks = 100)


dp.df %>%
  ggplot() +
  #geom_violin(aes(sample, depth)) +
  geom_boxplot(aes(reorder(sample, med_dp, decreasing = T), depth, fill = med_dp<6 ), outliers = FALSE) +
  #geom_point(aes(sample, depth), alpha = 0) +
  scale_y_continuous(name = "Depth") +
  scale_fill_manual(values = c(col_pal_mechanitis[5], "orange")) +
  coord_flip() +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank())


dp.df %>% filter(med_dp<10) %>%
  ggplot() +
  #geom_violin(aes(sample, depth)) +
  geom_boxplot(aes(reorder(sample, med_dp, decreasing = T), depth, fill = med_dp<6 ), outliers = FALSE) +
  #geom_point(aes(sample, depth), alpha = 0) +
  scale_y_continuous(#position = "right",
                     #limits = c(0,50),
                     name = "Depth, samples DP<10") +
  scale_fill_manual(values = c(col_pal_mechanitis[5], "orange")) +
  coord_flip() +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank())

# if many ind this is not informative
# d <- ggExtra::ggMarginal(size = 8,
#                     d, type = "density", 
#                     margins = "x",
#                     fill = col_pal_mechanitis[5],
#                     alpha = 0.5)


write.csv((dp.df %>% select(sample, med_dp) %>% distinct()), 
          paste(OUTPUT_DIR, "/tables/median_depth_per_sample.txt", sep = ""),
          quote = F, row.names = F)
summary(dp.df$depth)
print("Samples to potentially exclude with low depth (<6):")
dp.df %>% filter(med_dp<6) %>% select(sample, med_dp) %>% distinct()

```

```{r idepth_normalised, echo=FALSE}

head(dp.df)

auto_medians <- aggregate(
    depth ~ sample,
    data = dp.df[!grepl("^[ZW]", dp.df$CHROM), ],
    FUN = median,
    na.rm = TRUE
)

dp_norm.df <- merge(dp.df, auto_medians, by = "sample",
            suffixes = c("", "_auto_median"))

dp_norm.df <- dp_norm.df %>% 
  group_by(sample, CHROM) %>%
  mutate(med_dp_chr=median(depth, na.rm = T)) %>%
  mutate(depth_norm = med_dp_chr/depth_auto_median)

dp_norm.df %>% select(sample, depth_norm, CHROM) %>% distinct() %>%
  ggplot() +
  geom_point(aes(sample, depth_norm), colour = col_pal_mechanitis[5]) +
  scale_y_continuous(name = "Depth (normalised)") +
  scale_y_continuous(name = paste("Sample (", length(dp_norm.df$sample %>% unique()), " individuals)", sep = "")) +
  #scale_fill_manual(values = c(rep(col_pal_mechanitis, 6))) +
  #coord_flip() +
  facet_wrap(~CHROM) +
  theme_classic() +
  theme(axis.text.x = element_blank())



#select potential males
threshold <- 0.75

high_dp_samples <- dp_norm.df %>% select(sample, depth_norm, CHROM) %>% distinct() %>%
  filter(CHROM %in% c("Z", "Z1")) %>%
  filter(depth_norm > threshold)

print("Number of potential males, samples with normalised coverage > 0.75:")
length(high_dp_samples$sample)

write.csv(high_dp_samples, paste(OUTPUT_DIR, "/tables/potential_males.csv", sep = ""), quote = FALSE, 
            row.names = F)



```


### Individual missingness

```{r miss_ind, echo=FALSE}

indMiss <- apply(gt_mat, MARGIN = 2, function(x){ sum(is.na(x)) })
indMiss <- indMiss/nrow(vcf)

indMiss.df <- as.data.frame(indMiss)
indMiss.df$sample <- rownames(indMiss.df)

summary(indMiss.df)

p <- indMiss.df %>%
  ggplot() +
  geom_bar(aes(reorder(sample, indMiss, decreasing = T),indMiss*100), fill = "gold3", colour = "grey",  stat="identity") +
  geom_point(aes(reorder(sample, indMiss, decreasing = T),indMiss*100) , alpha = 0) +
  scale_y_continuous(#position = "right",
                     #limits = c(0,50),
                     name = "Missingness (%)") +
  #scale_fill_manual(values = rep(col_pal_mechanitis, 7)) +
  coord_flip() +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "grey"), 
        axis.title.y = element_blank(),
        legend.position = "None")


p <- ggExtra::ggMarginal(p, type = "density", 
                    margins = "x",
                    fill = "gold3",
                    alpha = 0.5,
                    size = 8)



indMiss.df %>% filter(indMiss > 0.05) %>%
  ggplot() +
  geom_bar(aes(reorder(sample, indMiss, decreasing = T),indMiss*100), fill = "gold3", colour = "grey",  stat="identity") +
  geom_point(aes(reorder(sample, indMiss, decreasing = T),indMiss*100) , alpha = 0) +
  scale_y_continuous(#position = "right",
                     #limits = c(0,50),
                     name = "Missingness (%), > 0.05",
                     breaks = seq(10, 100, 10)) +
  #scale_fill_manual(values = rep(col_pal_mechanitis, 7)) +
  coord_flip() +
  theme_classic() +
  theme(panel.grid.major.x = element_line(colour = "darkgrey"),
        axis.title.y = element_blank(),
        legend.position = "None")

print("Potential individuals to remove:")
indMiss.df %>% filter(indMiss > 0.2)

```



### Heterozygosity

```{r heterozygosity, echo=FALSE}

# Expected heterozygosity
gt <- extract.gt(vcf, element = "GT")
He_MAF_EHET <- compute_He_MAF_EHET(gt)


# Observed heterozygosity
gt <- extract.gt(vcf, element = "GT")
O_HET <- colSums(is.het(gt, na_is_false = TRUE))
n_called_sample <- colSums(!is.na(gt))
Het_prop <- O_HET / n_called_sample

H_prop.df <- data.frame(
  sample = names(Het_prop),
 Het_prop = Het_prop
)

ggplot(H_prop.df, aes(x = sample, y = Het_prop)) +
  geom_point(fill = col_pal_mechanitis[1], shape = 21, colour="grey") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "None")


# Inbreeding coefficient
F_s <- 1 - (O_HET / He_MAF_EHET$E_HET)
F_s

summary(F_s)

df_F <- data.frame(
  sample = names(F_s),
 F_s = F_s
)

ggplot(df_F, aes(x = sample, y = F_s)) +
  geom_col(fill = col_pal_mechanitis[1]) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "None")

ggplot(df_F, aes(x = F_s)) +
  geom_histogram(binwidth = 0.01) +
  theme_classic()



```


Population structure
If you have multiple populations, F can be high because of the Wahlund effect:
Mixing populations with different allele frequencies → apparent homozygosity excess.

F can be artificially inflated (more homozygotes due to allele dropout) if your VCF has:
low coverage,lots of missing data, poor SNP calling

These are rules of thumb, not strict cutoffs:
F	Likely meaning
0 to 0.05	normal
0.05 to 0.15	mild inbreeding or structure
0.15 to 0.30	moderate inbreeding
> 0.30	high inbreeding or serious data issues

Summary
F = 0 → random mating (HWE)
F > 0 → more homozygotes than expected (inbreeding / structure)
F < 0 → more heterozygotes than expected (outbreeding / admixture)



### PCA

```{r pca, echo=FALSE}
library(SNPRelate)


VCF_FILE=paste(WD, "/vcf_thinned/vcf_thin_all_snps.vcf.gz", sep = "")
META="/lustre/scratch125/tol/teams/meier/users/kn9/recombination/rawdata_to_vcf/metadata/formated_metadata_all.csv"

vcf <- read.vcfR(VCF_FILE)
vcf

# rename CHROM
vcf@fix[, "CHROM"] <- sub("^SUPER_", "", vcf@fix[, "CHROM"])

VCF_FILE_renamed=paste(WD, "/vcf_thinned/vcf_thin_all_snps_renamed.vcf.gz", sep = "")

# write a new VCF
write.vcf(vcf, file = VCF_FILE_renamed)

# convert renamed VCF to GDS format
vcf.fn <- VCF_FILE_renamed
gds.fn <- paste(OUTPUT_DIR,"/tables/gds_", Sys.Date(), ".fn", sep = "")

snpgdsVCF2GDS(vcf.fn, gds.fn, method="biallelic.only")
snpgdsSummary(gds.fn)

# open GDS
genofile <- snpgdsOpen(gds.fn)


# LD-pruning
snpset <- snpgdsLDpruning(genofile, ld.threshold=0.2)
str(snpset)

# Get all selected snp id
snpset.id <- unlist(unname(snpset))
head(snpset.id)

# run PCA
pca <- snpgdsPCA(genofile, snp.id=snpset.id, num.thread=2)

# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))

# make a data.frame
tab <- data.frame(sample.id = pca$sample.id,
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    stringsAsFactors = FALSE)
head(tab)

tab$sample.id <- sub("_Mechanitis_polymnia", "", tab$sample.id)
tab$sample.id <- sub("^[0-9]+-", "", tab$sample.id)

metadata <- read.csv(META, header = TRUE, sep = ",", na.strings = c("NA", ""))[,1:4]
#colnames(metadata) <- c("sample.id", "species", "subspecies", "Country")
metadata$Country <- toupper(metadata$Country)


tab2 <- left_join(tab, metadata, by = c("sample.id"="CAMID"))


# inspired from CAM0707355
col_pal_mechanitis <- c("orange3","#8A5628", "gold2", "#2D1A12","beige", "#9E8757","#D6D5D6", "white")

tab2 %>% filter(!is.na(Country)) %>%
  ggplot() +
  geom_point(aes(EV1, EV2,
                 fill = Country), size = 3, shape=21, colour = "black") +
  scale_fill_manual(values = c(col_pal_mechanitis, "purple", "")) +
  scale_y_continuous(paste("PC2 (",round(pc.percent[2], 2), ")", sep = "")) +
  scale_x_continuous(paste("PC1 (",round(pc.percent[1], 2), ")", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 20))


  
ggsave(filename=paste(OUTPUT_DIR,"/plots/pca.pdf", sep = ""),
       height = 4,
       width = 6
       )

tab2 %>% filter(!is.na(Country)) %>%
  ggplot() +
  geom_point(aes(EV1, EV2,
                 fill = subspecies), size = 3, shape = 21, colour = "black") +
  scale_fill_manual(values = c(col_pal_mechanitis, "purple", "blue")) +
  scale_y_continuous(paste("PC2 (",round(pc.percent[2], 2), ")", sep = "")) +
  scale_x_continuous(paste("PC1 (",round(pc.percent[1], 2), ")", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 20))

ggsave(filename=paste(OUTPUT_DIR,"/plots/pca_subsp.pdf", sep = ""),
       height = 4,
       width = 6
       )

# close
snpgdsClose(genofile)

```

Columbia and Ecuador in one cluster (proceriformis, dorissides, NA), Panama, close by (chimborazona, caucaensis,isthmia). 
Peruvian, French Guiana and Surinam (proceriformis (second cluster), polymnia, and NA) in one cluster
Brazil (casabranca) close to Peruvian cluster (South Peru, eurydice)

